---
title: "Metagenomic visualization"
author: "Hong Yang (hong.yang@kcl.ac.uk)"
date: "03/11/2022"
---

In this workshop, we will learn how to visualize data using R. we will use the output generated in previous sessions, including abundance data from quantification and results statistical analysis. 

### 1. what kinds of plot we can find in a metagenomic-associated paper?
We will look over several papers,please refer to the html file in the same folder.

#### Boxplot
#### Heatmap

### 2. Load necessary packages for this session

```{r, message=FALSE}
## A list of packages that we will use for this session
pkgs = c("tidyverse","ggplot2","phyloseq","ggpubr","stringr","dutchmasters","ggsci","devtools")

## Install the package if it has't been installed yet
for (i in pkgs){
  if(! i %in% installed.packages()){
    BiocManager::install(i, ask = F, update = F)
  }
}
## Load all the packages
invisible(lapply(pkgs, function(x) library(x, character.only=TRUE)))

## Alterative way for smplot
devtools::install_github('smin95/smplot')
library(smplot)
## remove the intermediate variables
rm(pkgs, i)
```


### 3. Load metageneomics data and preprocess data
#### 3.1 Load phyloseq object and results from statistical analysis
```{r, message=FALSE}
## phyloseq object that you generated
load('phyloseq_object_relative_abundance_2022.Rdata')
## Results from statistical analysis
load('Alpha_diversity_by_read.Rdata')
```

- Data structure of phyloseq object

```{r}
phylo_obj
```
- sample data
```{r}
phylo_obj@sam_data[1:3,1:5]
```
- otu table
```{r}
phylo_obj@otu_table[1:3,1:3]
```
- tax table
```{r}
phylo_obj@tax_table[1:3,1:3]
```

#### 3.2 Preprocessing, filter OTUs based on abundance
- Why?

```{r}
pheatmap::pheatmap(phylo_obj@otu_table)
```

- How?

```{r}
#e.g., only OTUs with a mean greater than 10^-5 are kept.
phylo_obj_filter = filter_taxa(phylo_obj, function(x) mean(x) > 1e-5, TRUE)
```

- Difference?

```{r}
#number of OTUs in orignial phyloseq object.
dim(phylo_obj@otu_table)
```

```{r}
#number of OTUs after filtering
dim(phylo_obj_filter@otu_table)
```


#### 3.3 Preprocessing, subset most abundant OTUs

```{r}
total = median(sample_sums(phylo_obj_filter))
abundant_OTU = filter_taxa(phylo_obj_filter, function(x) sum(x > total*0.20) > 0, TRUE)
abundant_OTU
```


### 4. Visualization
#### 4.1 data transition for plotting
- use psmelt() function to combine three data tables in phyloseq object

```{r}
pd = psmelt(phylo_obj_filter)
pd_abundant = psmelt(abundant_OTU)
head(pd)
```

6762 obs. of 15 variables

- formatting step for data of abundance
```{r}
pd = pd %>% mutate(Kingdom = str_replace(Kingdom, 'k__',''),
                   Phylum = str_replace(Phylum, 'p__',''),
                   Class = str_replace(Class, 'c__',''),
                   Order = str_replace(Order, 'o__',''),
                   Family = str_replace(Family, 'f__',''),
                   Genus = str_replace(Genus, 'g__',''),
                   Species = str_replace(Species, 's__',''))
pd[1:3,9:12]
```

-  formatting step for results from statistical analysis
```{r}
head(Adiv)
```

- Add metadata to the table
```{r}
Adiv_meta = Adiv %>% mutate(SampleID = rownames(.)) %>%
  right_join(SampleData, by = 'SampleID') %>%
  select(SampleID, everything())
head(Adiv_meta)
```

- Check the column name of combined data.frame
```{r}
colnames(Adiv_meta)
```

#### 4.2 Creat graph in R 
- ggplot + geom_?() functions
```{r}
## Create a basic layer by setting up the x- and y-axis of graph
p1 = ggplot(Adiv_meta, aes(x = SampleID, y = BMI))

## Add the second layer, like barplot 
p2 = ggplot(Adiv_meta, aes(x = SampleID, y = BMI)) + geom_bar(stat = 'identity')

## Add the second layer, like barplot (side-by-side)
p3 = ggplot(Adiv_meta, aes(x = SampleID, y = BMI, fill = Gender)) + 
  geom_bar(stat = 'identity', position = position_dodge())

## Add the second layer, like barplot (stacked)
p4 = ggplot(Adiv_meta, aes(x = Condition, y = BMI, fill = Gender)) + 
  geom_bar(stat = 'identity',position = "fill")

## Add the second layer, like boxplot
p5 = ggplot(Adiv_meta, aes(x = Condition, y = BMI)) + 
  geom_boxplot()

## Add the second layer, like heatmap
p6 = ggplot(Adiv_meta, aes(x = SampleID, y = '')) + 
  geom_tile(aes(fill = BMI))

ggarrange(p1,p2,p3,p4,p5,p6, labels = c("p1", "p2", "p3", "p4","p5","p6"))
```

##### [summary of Geoms in ggplot] 
- ggplot() + geom_bar(), create a bar plot
- ggplot() + geom_violin(), create a violin plot
- ggplot() + geom_boxplot(), create a boxplot
- ggplot() + geom_tile(), create a heatnmap
...

[Check more Geoms](https://ggplot2.tidyverse.org/reference/#layers) 

##### [exercise] 
- create a violin plot and shows the difference of BMI between Conditions
```{r}
# Add your code below:

```

Q: how to improve these graphs, in case:

- change the color?
- fix the crowded x-axis labels?
- add a title to my graph?
- remove title in x.asix and/or y.axis?
...

you name it!

#### 4.3 Graph theme, including default theme and alternative theme
- Components of the graph and corresponding functions that can be used
```{r}
### create a graph using default theme
p = ggplot(Adiv_meta, aes(x = Condition, y = BMI)) + 
  geom_boxplot() + 
  ggtitle('default theme')

## add the third layer with customized theme 
## change the theme and check the difference between these existed themes
p1 = p + theme_bw() + ggtitle('theme_bw()')
p2 = p + theme_linedraw() + ggtitle('theme_linedraw()')
p3 = p + theme_classic() + ggtitle('theme_classic()')
p4 = p + theme_dark() + ggtitle('theme_dark()')
p5 = p + theme_minimal() + ggtitle('theme_minimal()')

### arrange above plots into one and display
ggarrange(p,p1,p2,p3,p4,p5, labels = c("p", "p1","p2","p3","p4","p5"))
```

#### 4.4 Modify theme as you prefer

To modify an individual theme component you use code like plot + theme(element.name = element_function()). 

1. element.name including axis.title, axis.title.x (axis.title.y), axis.text.x (or axis.text.y),legend.position, ect. 
2. element_function() including element_text(), element_blank()
- element_text(), will show the text in the graph
- element_blank(), will remove the text in the graph

- [Modify components of a theme](https://ggplot2.tidyverse.org/reference/theme.html)

For example:
```{r}
p = ggplot(Adiv_meta, aes(x = Condition, y = BMI)) + 
  geom_boxplot()  +
  theme_classic() +
  ggtitle('theme_classic()')

# add argument inside of geom_boxplot() function, e.g, fill and width
p1 = ggplot(Adiv_meta, aes(x = Condition, y = BMI)) + 
  geom_boxplot(aes(fill = Condition))  +
  theme_classic() +
  ggtitle('change width and color of box')

# in case, the length of text is too long, we can rotate the label. 
# change the color of title and text in x-axis
# customize this argument within theme function: axis.title.x or axis.title.y 
# for example:
p2 = ggplot(Adiv_meta, aes(x = Condition, y = BMI)) + 
  geom_boxplot(aes(fill = Condition))  +
  theme_classic() +
  ggtitle('change color of x.title') +
  theme(axis.title.x = element_text(colour = 'red'))

# Multiple modification
# Modify the graph by
# change the color of title in x-axis
# rotate x-axis text
# change the legend position and

p3 = ggplot(Adiv_meta, aes(x = Condition, y = BMI)) + 
  geom_boxplot(aes(fill = Condition))  +
  theme_classic() +
  ggtitle('after mutiple modification') +
  theme(axis.title.x = element_text(colour = 'red'),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5, colour = 'red'),
        legend.position = 'left') # can be in left, bottom, top, or none

ggarrange(p, p1, p2, p3, labels = c("p", "p1", "p2","p3"))
```

##### The relative abundance of bacterial at the Phylum level in each sample

- check all the column names of pd that contains abundance

```{r}
colnames(pd)
```

```{r}
ggplot(data = pd, aes(x = Sample, y = Abundance, fill = Class)) + 
  geom_bar(stat = 'identity',position = "fill")
```


#### [exercise]:

how to solve the staked x-axis label?

- use theme() function to modify it. 

```{r}
# add your code here:


```

#### [exercise]:

how to change the float to percentage on y-axis label?

- use scale_y_continuous() function to modify it. 

```{r}
# add your code here:

```

##### how to add condition information of each sample
- use facet_grid() or facet_wrap()
- check the usage of these function using ?function in console.

```{r}
ggplot(data = pd, aes(x = Gender, y = Abundance)) + 
  geom_bar(stat = 'identity',position = "fill", aes(fill = Class)) +
  theme_classic2() +
  facet_grid(. ~ Condition, scales="free")
```


#### 4.5 Modify the color of bar/box
- use scale_fill_manual() function
- A color can be specified either by name (e.g.: “red”) or by hexadecimal code (e.g. : “#FF1234”) 
- please keep in mind that the number of colors you assign on your data should be same as the number of groups

```{r}
length(unique(pd$Class))
```
So, we need to supply 15 different colors to the corresponding bacterial at class level.
- [colors in R](http://www.sthda.com/english/wiki/colors-in-r)

```{r}
ggplot(data = pd, aes(x = Sample, y = Abundance)) + 
  geom_bar(stat = 'identity',position = "fill", aes(fill = Class), width = 2) +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6)) +
  facet_wrap(. ~ Condition, scales="free") +
  scale_fill_manual(values = terrain.colors(15)) 
# you can also use rainbow(15) or heat.colors(15) or topo.colors(15) or terrain.colors(15)  
```

- you can also other color systems from package, e.g, defined color palette by the scientific journals when the number of colors you need less than 10 (category)
- for example use scale_fill_npg() or  scale_fill_aaas() or scale_fill_lancet() or [more](https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html) in ggsci package


```{r}
#length(unique(pd$Phylum))
#6
p = ggplot(data = pd, aes(x = Condition, y = Abundance)) + 
  geom_bar(stat = 'identity', position = "fill", aes(fill = Phylum),width = 0.8) + 
  theme_classic() +
  scale_fill_simpsons()

p1 = ggplot(data = pd, aes(x = Condition, y = Abundance)) + 
  geom_bar(stat = 'identity', position = "fill", aes(fill = Phylum),width = 0.8) + 
  theme_classic() +
  scale_fill_uchicago()

ggarrange(p,p1, nrow = 2)
```

#### [exercise]: 

Create a side-by-side barplot showing the relative abundance of bacterial at Kingdom level and modify the color, which are commonly used in the paper published in Lancet journals.

```{r}
# add your code below:

```

#### 4.6 Adding statistical results on a graph



- use stat_compare_means() in ggpubr package for statistical analysis

```{r}
p1 = ggplot(data = Adiv_meta, aes(x = Condition, y = Shannon)) + 
  geom_boxplot(aes(fill = Condition)) +
  geom_jitter() +
  theme_classic()+
  scale_fill_jco() +
  stat_compare_means(method = "wilcox.test") +
  scale_y_continuous(expand = c(0.01, 0.2))


p2 = ggplot(data = Adiv_meta, aes(x = Condition, y = Shannon)) + 
  geom_violin(aes(fill = Condition)) +
  theme_classic()+
  scale_fill_jco() +
  stat_compare_means(method = "wilcox.test") +
  scale_y_continuous(expand = c(0.01, 0.2))

p3 = ggplot(data = Adiv_meta, aes(x = Condition, y = Chao1)) + 
  geom_boxplot(aes(fill = Condition)) +
  theme_classic()+
  scale_fill_uchicago() +
  stat_compare_means(method = "wilcox.test",aes(label = ..p.signif..)) + 
  scale_y_continuous(expand = c(0.01, 20))


p4 = ggplot(data = Adiv_meta, aes(x = Condition, y = Chao1)) + 
  geom_violin(aes(fill = Condition)) +
  theme_bw()+
  facet_grid(.~ Gender) +
  scale_fill_uchicago() +
  stat_compare_means(method = "wilcox.test",aes(label = ..p.signif..)) +
  scale_y_continuous(expand = c(0.01, 20))


ggarrange(p1, p2,p3,p4, labels = c("p1","p2","p3","p4"), nrow = 2, ncol = 2)
```

- Student T-test or wilcox.test?

#### [exercise]: 

How to evaluate the difference of Shannon/Chao1 between Gender in each condition, significant or not?
```{r}
# add your code below:

```


#### 4.7 Correlation

- how to perform the correlation analysis using R?
```{r}
cor.test(Adiv_meta$BMI, Adiv_meta$Age, method = 'spearman')
```
- are BMI and Shannon index significantly correlated?

```{r}
# add your code below:
cor.test(Adiv_meta$BMI, Adiv_meta$Shannon, method = 'spearman')
```

```{r}
p1 = ggplot(data = Adiv_meta, aes(x = BMI, y = Age)) +
  geom_point(shape = 21, fill = '#0f993d', color = 'white', size = 3) +
  sm_corr_theme() + 
  sm_statCorr(corr_method = "spearman")

p2 = ggplot(data = Adiv_meta, aes(x = BMI, y = Shannon)) +
  geom_point(shape = 21, fill = 'purple', color = 'white', size = 3) +
  sm_corr_theme() + 
  sm_statCorr(corr_method = "spearman")

ggarrange(p1, p2, labels = c("p1","p2"))
```


#### [excerise]:

how about the correlation between Age and Shannon index

```{r}
# Add your code below:
```

#### 4.8 Heatmap showing the most abundant OTUs across samples
- use geom_tile() in ggplot2

```{r}
pd_abundant = pd_abundant %>% mutate(Kingdom = str_replace(Kingdom, 'k__',''),
                   Phylum = str_replace(Phylum, 'p__',''),
                   Class = str_replace(Class, 'c__',''),
                   Order = str_replace(Order, 'o__',''),
                   Family = str_replace(Family, 'f__',''),
                   Genus = str_replace(Genus, 'g__',''),
                   Species = str_replace(Species, 's__',''))

ggplot(data = pd_abundant, aes(x=SampleID, y=Species, fill = log10(Abundance+1))) +
  geom_tile() +
  theme_classic() +
  facet_grid(.~ Condition, scales = 'free') +
  scale_fill_gradient2(midpoint = 0.8, mid = 'white',low = '#131D5C', high = '#CE3633') + 
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank()) +
  rremove('x.ticks')
```

- it is more flexible using ggplot.

#### 4.9 Saving the graph in a file
- save your graph as a image, or as a pdf file using ggsave() functions
```{r}
# plotting and saving as a object in R (e.g, fig_a)
fig_a = ggplot(Adiv_meta, aes(x = Condition, y = BMI)) + 
  geom_boxplot(aes(fill = Condition)) +
  geom_jitter() +
  theme_bw() +
  ggtitle('BMI in different condition')

# at least 4 arguments should be given to ggsave() function
# first, a object that represents the graph you generate, (e.g., fig_a in the case)
# second, file name and the folder you will save the file
# third and forth are about width and height of graph, respectively
ggsave(fig_a, filename = 'Figures/00.BMI.condition.pdf', width = 4, height = 3)
# you can save the graph in other version
# ggsave() support : "eps", "ps", "tex" (pictex), "pdf", "jpeg", "tiff", "png", "bmp", "svg" 
```

#### 4.10 Organize multiple graphs into one and save
```{r}
fig_b = ggplot(data = pd_abundant, aes(x=SampleID, y=Species, fill = log10(Abundance+1))) +
  geom_tile() +
  theme_classic() +
  facet_grid(.~ Condition, scales = 'free') +
  scale_fill_gradient2(midpoint = 0.8, mid = 'white',low = '#131D5C', high = '#CE3633') + 
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(), 
        legend.position = 'bottom') +
  rremove('x.ticks') +
  ggtitle('Abundant species across samples')

combined_fig = ggarrange(fig_a, fig_b, labels = c("A","B"))
ggsave(combined_fig, filename = 'Figures/00.combined_fig.pdf', width = 10, height = 5)

```



-End, thank you!!



